---
title: "OTN and DFO Activity Summary in Eastern Canada: DRAFT"
author: Naomi Tress and Jon Pye (OTN), Heather Bowlby, Rémi Daigle, and Catalina Gomez (DFO)
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    code_folding: hide
---

## About this Document 

- This summary has been created using reproducible code (https://github.com/dfo-mar-odis/OTN-DFO-Summary), which gathers and filters data from the Ocean Tracking Network (OTN) to highlight activities in which Fisheries and Oceans Canada (DFO) has actively participated in eastern Canada.

- This summary is being prepared with the support of the OTN International Data Management Committee and DFO staff to provide an overview of current OTN DFO activities.

- Please note that while we strive to include all relevant DFO-OTN information in this summary, there may be instances where some projects are not listed. If you are a DFO investigator with an OTN project that you do not see here, we encourage you to contact OTN directly.

## DFO & OTN 

- The OTN is a global aquatic research, data management and partnership platform. Since 2008, OTN has been deploying state-of-the-art ocean monitoring equipment and marine autonomous vehicles (gliders) in key ocean locations and inland waters around the world. OTN’s technical capabilities expanded in 2020 with the addition of remotely operated vehicles (ROVs) and side scan sonar systems. Researchers around the world are using OTN’s global infrastructure and analytical tools to document the movements and survival of aquatic animals in the context of changing ocean and freshwater environments. For more information, please visit the [OTN site](http://oceantrackingnetwork.org/).

- [DFO](https://www.dfo-mpo.gc.ca/science/mammals-mammiferes/otn/index-eng.html) is a partner in the OTN effort. DFO researchers initiate projects, while the OTN provides support. OTN assumes a leading role in managing infrastructure and its maintenance, as well as providing equipment to support research projects. Some DFO scientists upload data to OTN's database to manage and ensure data quality.

- Both OTN and DFO are committed to the principles of open data, actively working to ensure transparency and accessibility in sharing data related to marine research and monitoring.

## OTN Data Wrangling

In this section, we gather data from OTN and leverage it to create informative maps and plots for this summary document. Our process involves data acquisition, filtering, and visualization. Below, we provide a brief overview of these steps, and you can expand the code sections to see the full data wrangling process.


```{r setup, include=FALSE}
library(lubridate)
library(leaflet)
library(plotly)
library(arcpullr)
library(sf)
library(tidyverse)
```

```{r Set project bounds}
proj_start <- ymd("20190101")
proj_end <- ymd("20230909")
proj_long_upp <- -40.00
proj_long_low <- -70.00
proj_lat_upp <- 60.00
proj_lat_low <- 40.00
```

```{r Import MPAs from Open Data, warning=FALSE, message=FALSE}
url <- "https://gisp.dfo-mpo.gc.ca/arcgis/rest/services/FGP/Oceans_Act_Marine_Protected_Areas/MapServer/0"
MPAs <- get_spatial_layer(url) %>% 
  st_make_valid() %>% 
  st_crop(xmin=proj_long_low,
          ymin=proj_lat_low,
          xmax=proj_long_upp,
          ymax=proj_lat_upp)
```


```{r Import receiver data from the geoserver, warning=FALSE, message=FALSE}
geoserver_receivers <- readr::read_csv('https://members.oceantrack.org/geoserver/otn/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=otn:stations_receivers&outputFormat=csv', guess_max = 13579)
```

```{r Import animal data from the geoserver, warning=FALSE, message=FALSE}
geoserver_tag_releases <- readr::read_csv('https://members.oceantrack.org/geoserver/otn/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=otn:animals&outputFormat=csv', guess_max = 13579)

geoserver_tag_releases <- geoserver_tag_releases %>%
  filter(yearcollected > 2018)
```

```{r Import project data from the geoserver, warning=FALSE, message=FALSE}
geoserver_projects <- readr::read_csv('https://members.oceantrack.org/geoserver/otn/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=otn:otn_resources_metadata_points&outputFormat=csv', guess_max = 13579)
```

```{r Filter receivers data}
# Filter the geoserver_receivers dataset to select relevant stations 

otn_stations <- geoserver_receivers %>%
  # Remove rows with missing deploy_date values
  filter(!is.na(deploy_date)) %>%
  
  # Select stations that fall within the project timeframe defined above
  filter((deploy_date > proj_start & deploy_date < proj_end) |
           (recovery_date < proj_end & recovery_date > proj_start) |
           (deploy_date < proj_end & is.na(recovery_date) & deploy_date > proj_start - duration(18, 'months')) |
           
           # Select specific models within certain date ranges
           (grepl('VR3', model) & deploy_date < proj_end & is.na(recovery_date) & deploy_date > proj_start - duration(4, 'years')) |
           (grepl('VR4', model) & deploy_date < proj_end & is.na(recovery_date) & deploy_date > proj_start - duration(6, 'years'))) %>%
  
  # Filter stations based on latitude and longitude bounds
  filter(stn_lat >= proj_lat_low & stn_lat <= proj_lat_upp &
           stn_long >= proj_long_low & stn_long <= proj_long_upp)
```

```{r Filter animals data}
# Define the columns to keep
columns_to_keep <- c("FID", "institutioncode", "datacenter_reference", "scientificname", "vernacularname","longitude", "latitude", "basisofrecord", "yearcollected", "collector", "classname")

# Filter and select the desired columns
otn_animals <- geoserver_tag_releases %>%
  filter(longitude >= proj_long_low & longitude <= proj_long_upp &
           latitude >= proj_lat_low & latitude <= proj_lat_upp) %>%
  select(all_of(columns_to_keep))
```

```{r Filter project data}
# Define the columns to keep
columns_to_keep_projects <- c("FID", "resource_full_name", "ocean", "seriescode", "status", "collaborationtype")

# Filter and select the desired columns
otn_projects <- geoserver_projects %>%
  filter(seriescode == "DFOCanada") %>%
  select(all_of(columns_to_keep_projects))
```

## OTN Summary of Receivers in Eastern Canada 

```{r OTN receivers map}
# Create a Leaflet map for receiver stations
map <- leaflet(otn_stations) %>%
  # Add default basemap
  addTiles() %>%
  # Add MPA polygons
  addPolygons(data = MPAs, 
              color = "yellow", 
              weight = 0, 
              fillOpacity = 1, 
              label = ~NAME_E) %>%
  
  # Add circle markers for each station
  addCircleMarkers(
    lng = ~stn_long,
    lat = ~stn_lat,
    label = ~paste(seriescode, instrumenttype, sep = " - "),
    radius = 0.5,  # Adjust the circle marker size 
    color = ~ifelse(seriescode == "DFOCanada", "red", "blue"),  # Marker color
    fillOpacity = 0.8  # Adjust fill opacity
  ) %>%
  
  # Add bounding box based on filter
  addRectangles(
    lng1 = proj_long_low,      # Left longitude
    lat1 = proj_lat_low,       # Lower latitude
    lng2 = proj_long_upp,      # Right longitude
    lat2 = proj_lat_upp,       # Upper latitude
    weight = 2.5,              # Border width
    color = "darkorange",      # Border color
    fill = FALSE               # Don't fill the rectangle
  ) %>%
  
  # Add legend for all elements of the plot
  addLegend(
    position = "bottomleft",
    colors = c("blue", "red", "darkorange", "yellow"),
    labels = c("All OTN", "OTN-DFO", "Data filter", "Marine Protected Areas"),
    title = "Legend"
  ) %>%
  
  # Add filter control displaying the range date
  addControl(
    html = sprintf("<strong>Filter:</strong> %s - %s", proj_start, proj_end)
  )
```

```{r OTN receivers table}
# Calculate the total number of receivers for each seriescode (OTN vs OTN-DFO).
summary_table <- otn_stations %>%
  group_by(seriescode) %>%
  summarize(
    Total_Receivers = n()
  ) %>%
  # Calculate the percentage relative to the total and round it.
  mutate(
    Percentage_Relative_To_Total = paste(round((Total_Receivers / sum(Total_Receivers)) * 100, 0), "%"))

# Rename columns
colnames(summary_table)[colnames(summary_table) == "Total_Receivers"] <- "Total # of Receivers"
colnames(summary_table)[colnames(summary_table) == "Percentage_Relative_To_Total"] <- "% Receivers relative to total"

# Print the summary table as a Markdown table.
table_receivers <- knitr::kable(
  summary_table, 
  format = "markdown",
  align = "l", 
  width = "100%"
)

```

_Disclaimer: Information includes data that has been filtered for specific criteria relevant for eastern Canada and does not encompass all available OTN data. For a comprehensive dataset, please visit the OTN website._

```{r receivers visuals and table}
# Display the map and table
map
table_receivers
```

## OTN Summary of Species Tagged in Eastern Canada as part of DFO projects 

This section is a summary of the OTN-DFO data, focusing on # of animals tagged by species in Eastern Canada. It includes the following elements:

- Number of animals tagged (i.e. number of tags deployed) displayed on the y-axis.
- Species displayed on the x-axis.
- A summary statistic comparing the number of animals tagged with involvement from DFO 

```{r barplot of animals tagged, warning=FALSE, message=FALSE}
# Create a summary table with counts of animals tagged by species.
summary_otn_animals <- otn_animals %>%
  group_by(vernacularname) %>%
  summarise(
    Count = n(),
    Count_DFO = sum(grepl("DFO", institutioncode))
  )

# Reorder the levels of vernacularname based on Count 
summary_otn_animals <- summary_otn_animals %>%
  arrange(desc(Count)) %>%
  mutate(vernacularname = factor(vernacularname, levels = vernacularname))

# Create a bar plot using ggplot2.
plot_animals <- ggplot(summary_otn_animals, aes(x = vernacularname)) +
  geom_bar(aes(y = Count, fill = "OTN"), stat = "identity", position = "dodge") +
  geom_bar(aes(y = Count_DFO, fill = "OTN-DFO"), stat = "identity", position = "dodge") +
  labs(
    title = "Number of Animals Tagged by Species in Eastern Canada",
    x = "Species",
    y = "Number of Animals Tagged"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip() +
  scale_fill_manual(values = c("OTN" = "blue", "OTN-DFO" = "red"))

# Convert the ggplot plot to an interactive plot using plotly.
interactive_plot <- ggplotly(plot_animals) %>%
  layout(legend = list(x = 0.5, y = 1))
```

```{r map of animals tagged}
# Create a Leaflet map to visualize were animals were originally tagged
map_animals <- leaflet(otn_animals) %>%
  addTiles() %>%
  
  # Add MPA polygons
  addPolygons(data = MPAs, 
              color = "yellow", 
              weight = 0, 
              fillOpacity = 1, 
              label = ~NAME_E) %>%

  # Add circle in the map for each animal tagged.
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    label = ~paste(institutioncode, scientificname, vernacularname, yearcollected, collector, sep = " - "),
    radius = 0.5,
    color = ~ifelse(grepl("DFO", institutioncode), "red", "blue"),
    fillOpacity = 0.8
  ) %>%

  # Add bounding box based on filter defined at the beginning of code.
  addRectangles(
    lng1 = proj_long_low,
    lat1 = proj_lat_low,
    lng2 = proj_long_upp,
    lat2 = proj_lat_upp,
    weight = 2.5,
    color = "darkorange",
    fill = FALSE
  ) %>%

  # Add legend for different elements of the map.
  addLegend(
    position = "bottomleft",
    colors = c("blue", "red", "darkorange", "yellow"),
    labels = c("All OTN", "OTN-DFO", "Data filter", "Marine Protected Areas"),
    title = "Legend"
  )

# Calculate the minimum and maximum years in the dataset.
min_year <- min(otn_animals$yearcollected)
max_year <- max(otn_animals$yearcollected)

# Add a text box to display the year collected range.
map_animals <- map_animals %>%
  addControl(
    html = sprintf("<strong>Year Collected Range:</strong> %d - %d", min_year, max_year),
    position = "topleft"
  )
```

```{r animals tagged visuals}
interactive_plot
map_animals
```

```{r animals tagged table}
# Calculate the percentage of OTN-DFO tagging for each species.
summary_otn_animals$Percentage = ifelse(summary_otn_animals$Count_DFO == 0, "0%", paste0(round(summary_otn_animals$Count / summary_otn_animals$Count_DFO * 100), "%"))

# Create a table with species, OTN total count, and % OTN-DFO tagging.
knitr::kable(
  summary_otn_animals[, c("vernacularname", "Count", "Percentage")],
  format = "markdown", 
  col.names = c("Species", "OTN Total", "% OTN-DFO Tagging"),
  caption = "Summary of Animals Tagged by Species in Eastern Canada"
)
```

## OTN and DFO Project Summaries 

This section provides an overview of the **`r nrow(otn_projects)`** DFO-OTN projects available in the OTN database. 

We have applied a basic filter in this section, considering projects affiliated with DFO, without specific filters related to years or geographic locations. 

Please note that while we strive to include all relevant DFO-OTN projects in this summary, there may be instances where some projects are not listed. If you are a DFO researcher with an OTN project that you do not see here, we encourage you to contact OTN directly.

For the most up-to-date and comprehensive information on OTN projects and their participants, please visit the official OTN website: [OTN Website](https://members.oceantrack.org/data/discovery/bycontact.htm)

```{r project visuals}
# Create a stacked bar plot of collaboration types by status using ggplot2.
stacked_bar_plot <- ggplot(otn_projects, aes(x = collaborationtype, fill = status)) +
  geom_bar() +  # Create a bar plot.
  labs(title = "Collaboration Type Distribution") +  # Set the plot title.
  theme_minimal() +  # Use a minimal theme.
  scale_fill_manual(values = c("ongoing" = "green", "completed" = "blue", "proposed" = "orange")) +  # Define fill colors.
  coord_flip()  # Flip the coordinates for horizontal bars.

# Convert the ggplot plot to a plotly object.
interactive_stacked_bar_plot <- ggplotly(stacked_bar_plot)

# Customize the layout of the interactive plot.
interactive_stacked_bar_plot %>%
  layout(legend = list(x = 0.5, y = 1))  # Adjust the position of the legend.
```

```{r project table}
# Create an HTML table using knitr::kable to display a summary of DFO-OTN projects.
knitr::kable(
  otn_projects[, c("resource_full_name", "ocean", "status", "collaborationtype")],
  caption = "Summary of DFO-OTN Projects",
  format = "html",
  col.names = c("Project Title", "Ocean", "Status", "Collaboration Type")
)
```

Future work includes creating the following additional information for projects:

- Project Title:
- Citation:
- Points of contact:
- Species tagged:
- Abstract:

## Questions and Suggestions

If you have any questions or suggestions regarding this reproducible report, please don't hesitate to reach out to Catalina.Gomez@dfo-mpo.gc.ca and/or Jonathan.Pye@dal.ca. We welcome any inquiries or ideas to enhance the content or clarity of this report.